---
title: "Fitting data to linear models"
output:
  html_document:
    df_print: paged
---

## Import packages

```{r echo=TRUE}
library(lme4)
library(car)
library(dplyr)
library(tidyr)
```

## Import and subset dataset
- `df_clean` contains all rows of data that does not have `Null`/`UNKWN` in `Prov` and `Con_ACT`
- `no_censor` contains all rows of df_clean where reason for discontinuation is not due to `censor`

```{r echo=FALSE}
require(tidyr)
require(dplyr)
df <- read.delim("../data/masterdf.txt")
df$Tx_duration <-as.numeric(df$Tx_duration)
df$Age <- ordered(df$Age, levels = c(levels(df$Age)))

df_clean <- df %>% drop_na(Prov, Con_ACT) %>% 
  mutate(Age_num = as.numeric(regmatches(Age, regexpr('[0-9]{2}', Age))))

no_censor <- df_clean %>% filter (reason != 'censor')
```


## `df_clean`
- Assume each row (patient) is independent

### 1. Test for homogeneity of variance
```{r echo=TRUE}
require(car)
#bartlett.test(Tx_duration ~ interaction(Con_ACT, Age_int, reason ,Prov, Sex), data=df_clean) 
#Error in bartlett.test.default(c(12, 29, 5, 4, 27, 13, 11, 6, 4, 1, 6, : there must be at least 2 observations in each group

leveneTest(Tx_duration ~ Con_ACT * Age * reason  *  Prov * Sex, data = df_clean)
leveneTest(Tx_duration ~ Con_ACT * reason  *  Prov * Sex, data = df_clean)
#fligner.test(Tx_duration ~ Con_ACT * Age * reason  *  Prov * Sex, data = df_clean) #non-parametric
```

### 2. Test for homogeneity of covariance
```{r echo=TRUE}
require(DFA.CANCOR)
#homogeneity(data = df_clean, 
#            groups = c('Prov', 'Con_ACT', 'Sex', 'reason'),
#            variables = c('Tx_duration', 'Age_num'))

```

### 3. Linear model and model residuals
```{r echo=TRUE}
model <- lm(Tx_duration ~ Con_ACT * Age + reason  +  Prov + Sex, data = df_clean)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)

model <- lm(Tx_duration ~ Con_ACT * Age_num + reason  +  Prov + Sex, data = df_clean)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)


```

### 4. Model summary
```{r echo=TRUE}
model <- lm(Tx_duration ~ Con_ACT * Age + reason  +  Prov + Sex, data = df_clean)
summary(model)
print(rbind(c('AIC', AIC(model)),
            c('BIC', BIC(model)),
            c('logLik', logLik(model))))

model <- lm(Tx_duration ~ Con_ACT * Age_num + reason  +  Prov + Sex, data = df_clean)
summary(model)
print(rbind(c('AIC', AIC(model)),
            c('BIC', BIC(model)),
            c('logLik', logLik(model))))

```

### 5. Linear mixed model
```{r echo=FALSE}
model <- lmer(Tx_duration ~ Con_ACT + Age_num  +(1|reason)+ (1|Prov) + (1|Sex), REML= F,data = df_clean)
#par(mfrow=c(2, 3))
hist(resid(model))
plot(model)
summary(model)

model <- lmer(Tx_duration ~ Con_ACT + Age  +(1|reason) + (1|Prov) + (1|Sex), REML= F,data = df_clean)
#par(mfrow=c(2, 3))
hist(resid(model))
plot(model)
summary(model)

```


## `no_censor`

### 1. Test for homogeneity of variance
Nothing passed test for homogeneity of variance
```{r echo=TRUE}
require(car)
#bartlett.test(Tx_duration ~ interaction(Con_ACT, Age_num, Prov, Sex), data=no_censor) 
#Error in bartlett.test.default(c(12, 29, 5, 4, 27, 13, 11, 6, 4, 1, 6, : there must be at least 2 observations in each group
leveneTest(Tx_duration ~ Con_ACT * Age  *reason*  Prov * Sex, data =df_clean)
leveneTest(Tx_duration ~ Con_ACT * reason* Prov * Sex, data =df_clean)

leveneTest(Tx_duration ~ Con_ACT * Age  *  Prov * Sex, data = no_censor)
leveneTest(Tx_duration ~ Con_ACT * Prov * Sex, data = no_censor)
#fligner.test(Tx_duration ~ Con_ACT * Age_num * reason  *  Prov * Sex, data = df_clean) #non-parametric
```

### 2. Test for homogeneity of covariance
```{r echo=TRUE}
require(DFA.CANCOR)
#homogeneity(data = no_censor, 
#            groups = c('Prov', 'Con_ACT', 'Sex', 'reason'),
#            variables = c('Tx_duration', 'Age_num'))

```

### 3. Linear model and model residuals
```{r eval=FALSE, include=FALSE}
model <- lm(Tx_duration ~ Con_ACT * Age +  Prov + Sex, data = no_censor)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)

#Do not consider this model because including Age_num gives unequal variance among groups
#model <- lm(Tx_duration ~ Con_ACT * Age_num +  Prov + Sex, data = no_censor)
#shapiro.test(model$residuals)
#print(c("mean residuals:", mean(model$residuals)))
#print(c("median residuals:", median(model$residuals)))
#par(mfrow=c(2, 3))
#hist(model$residuals)
p#lot(model)


```

### 4. Model summary
```{r eval=FALSE, include=FALSE}
model <- lm(Tx_duration ~ Con_ACT * Age  +  Prov + Sex, data = no_censor)
summary(model)
print(rbind(c('AIC', AIC(model)),
            c('BIC', BIC(model)),
            c('logLik', logLik(model))))
#Do not consider this model because including Age_num gives unequal variance among groups
#model <- lm(Tx_duration ~ Con_ACT * Age_num + reason  +  Prov + Sex, data = no_censor)
#summary(model)


```

### 5. Linear mixed model
```{r eval=FALSE, include=FALSE}
model <- lmer(Tx_duration ~ Con_ACT *Age_num+ (1|Sex) +(1|reason)+ (1|Prov),
              REML = F, 
              data =df_clean)
#par(mfrow=c(1, 2))
hist(resid(model))
plot(model)
summary(model)

model <- lmer(Tx_duration ~ Con_ACT *Age+ (1|Sex) +(1|reason)+ (1|Prov),
              REML = F, 
              data =df_clean)
#par(mfrow=c(1, 2))
hist(resid(model))
plot(model)
summary(model)

model <- lmer(Tx_duration ~ Con_ACT *Age_num  + (1|Prov) + (1|Sex),REML = F, data = no_censor)
#par(mfrow=c(2, 3))
hist(resid(model))
plot(model)
summary(model)

```


## `scaled_df_clean`and `scaled_no_censor`
```{r echo = FALSE}

scale_this <- function(x){
  (x - mean(x, na.rm=TRUE)) / sd(x, na.rm=TRUE)
}


scaled_df_clean <- df_clean %>%
  mutate(Tx_duration = scale_this(Tx_duration),
         Age_num = scale_this(Age_num))

scaled_no_censor <- scaled_df_clean %>% filter (reason != 'censor')

```

### 1. Test for homogeneity of variance

Passed homogeneity of variance:
- `scaled_df_clean` with `Age_num` as numeric explanatory variable
- `scaled_df_clean` with `Age`
- `scaled_no_censor` with `Age_num`

```{r echo=TRUE}
require(car)
df <- scaled_df_clean
leveneTest(Tx_duration ~ Con_ACT *Age *  Prov * Sex, data = df)
leveneTest(Tx_duration ~ Con_ACT * Prov * Sex, data = df)
#bartlett.test(Tx_duration ~ interaction(Con_ACT, Age_num, Prov, Sex),
#              data=df) 


df <- scaled_no_censor
leveneTest(Tx_duration ~ Con_ACT *Age *  Prov * Sex, data = df)
leveneTest(Tx_duration ~ Con_ACT * Prov * Sex, data = df)
#fligner.test(Tx_duration ~ Con_ACT * Age_num * reason  *  Prov * Sex, data = df_clean) #non-parametric
```

### 2. Test for homogeneity of covariance
```{r eval=FALSE, include=FALSE}
require(rstatix)
require(covTestR)
test <- scaled_df_clean %>% 
  mutate(Group1 = as.factor(paste0(reason, Prov, Con_ACT, Sex)),
         Group2 = as.factor(paste0(Age, reason, Prov, Con_ACT, Sex))) %>% 
  select(Tx_duration, Age_num, Group1) 
#f <- box_m(data = test[,-3], group = test[,3])
#f1<-homogeneityCovariances(test, 
#            group = Group1)
f1<- test %>%  group_by(Group1) %>% 
      tally()
#print(f1)
```

### 3. Linear model and model residuals

#### With interaction
```{r echo=TRUE}

#`scaled_df_clean` with `Age_num` as numeric explanatory variable
df <- scaled_df_clean
model <- lm(Tx_duration ~ Con_ACT * Age_num + reason+ Prov + Sex, data = df)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)



#`scaled_df_clean` with `Age`
df <- scaled_df_clean
model <- lm(Tx_duration ~ Con_ACT * Age + reason+ Prov + Sex, data = df)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)


#`scaled_no_censor` with `Age_num`
df <- scaled_no_censor
model <- lm(Tx_duration ~ Con_ACT * Age_num + Prov + Sex, data = df)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)


```

#### Without interaction
```{r echo=TRUE}

#`scaled_df_clean` with `Age_num` as numeric explanatory variable
df <- scaled_df_clean
model <- lm(Tx_duration ~ Con_ACT + Age_num + reason+ Prov + Sex, data = df)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)



#`scaled_df_clean` with `Age`
df <- scaled_df_clean
model <- lm(Tx_duration ~ Con_ACT + Age + reason+ Prov + Sex, data = df)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)


#`scaled_no_censor` with `Age_num`
df <- scaled_no_censor
model <- lm(Tx_duration ~ Con_ACT + Age_num + Prov + Sex, data = df)
shapiro.test(model$residuals)
print(c("mean residuals:", mean(model$residuals)))
print(c("median residuals:", median(model$residuals)))
par(mfrow=c(2, 3))
hist(model$residuals)
plot(model)


```

### 4. Model summary
```{r echo=TRUE}
#`scaled_df_clean` with `Age_num` as numeric explanatory variable
df <- scaled_df_clean
model <- lm(Tx_duration ~ Con_ACT * Age_num + reason+ Prov + Sex, data = df)
summary(model)
print(rbind(c('AIC', AIC(model)),
            c('BIC', BIC(model)),
            c('logLik', logLik(model))))

#`scaled_df_clean` with `Age`
df <- scaled_df_clean
model <- lm(Tx_duration ~ Con_ACT * Age + reason+ Prov + Sex, data = df)
summary(model)
print(rbind(c('AIC', AIC(model)),
            c('BIC', BIC(model)),
            c('logLik', logLik(model))))

#`scaled_no_censor` with `Age_num`
df <- scaled_no_censor
model <- lm(Tx_duration ~ Con_ACT * Age_num + Prov + Sex, data = df)
summary(model)
print(rbind(c('AIC', AIC(model)),
            c('BIC', BIC(model)),
            c('logLik', logLik(model))))
#Do not consider this model because including Age_num gives unequal variance among groups
#model <- lm(Tx_duration ~ Con_ACT * Age_num + reason  +  Prov + Sex, data = no_censor)
#summary(model)


```

### 5. Linear mixed model

##### Preferred models
```{r echo=TRUE}
require(lmerTest)
model <- lmer(Tx_duration ~ Con_ACT + Age_num+  reason + Sex  + (1|Prov),REML = F, data = scaled_df_clean)
par(mfrow=c(1, 1))
hist(resid(model))
shapiro.test(resid(model))
qqnorm(y = resid(model))
plot(model)
summary(model)

model <- lmer(Tx_duration ~ Con_ACT + Age_num+  reason + (1|Sex)  + (1|Prov),REML = F, data = scaled_df_clean)
#par(mfrow=c(1, 1))
hist(resid(model))
shapiro.test(resid(model))
qqnorm(y = resid(model))
plot(model)
summary(model)

model <- lmer(Tx_duration ~ Con_ACT * Age_num+  reason + (1|Sex)  + (1|Prov),REML = F, data = scaled_df_clean)
#par(mfrow=c(1, 1))
hist(resid(model))
shapiro.test(resid(model))
qqnorm(y = resid(model))
plot(model)
summary(model)

```

##### Other models
```{r echo=TRUE}
model <- lmer(Tx_duration ~ Con_ACT * Age + reason + (1|Prov) + (1|Sex) , REML = F,  data = scaled_df_clean)
hist(resid(model))
qqnorm(y = resid(model))
plot(model)
shapiro.test(resid(model))
summary(model)
#vcov(model)


model <- lmer(Tx_duration ~ Con_ACT * Age_num  + (1|Prov) + (1|Sex), REML = F, data = scaled_df_clean)
hist(resid(model))
shapiro.test(resid(model))
plot(model)
qqnorm(y = resid(model))
summary(model)

model <- lmer(Tx_duration ~ Con_ACT * Age_num  + Sex + (1|Prov) , REML = F, data = scaled_df_clean)
hist(resid(model))
shapiro.test(resid(model))
plot(model)
qqnorm(y = resid(model))
summary(model)
```